---
title: "Nhanes data analysis"
output: html_document
date: "2023-02-17"
---
Nhanes data analysis 

```{r}
#Access the dataset in R and review the dataset documentation via the following commands
install.packages("NHANES_2.1.0.tgz", repos = NULL, type = "https://cran.rstudio.com/bin/macosx/big-sur-arm64/contrib/4.2/NHANES_2.1.0.tgz")
library(ggplot2)
library(NHANES)
data(NHANES)
```


Data exclusion 
```{r}
# excluding people who are under 21 years old, women who are pregnant, and people who have diabetes. 
older21 <- subset(NHANES, Age >= 21)

nopreganant <- subset(older21, PregnantNow == "No" | is.na(PregnantNow))

nodiabetes <- subset(nopreganant, Diabetes == "No" | is.na(Diabetes))

final_data <- nodiabetes
```

Descriptive Analysis
```{r}
summary(final_data)
```

```{r}
# distribution graph for the blood pressure 
# Create a histogram of the BPSysAve variable with white background and gray fill color
ggplot(NHANES, aes(x=BPSysAve)) +
  geom_histogram(binwidth=5, color="gray", fill="lightgray") +
  labs(title="Distribution of Systolic Blood Pressure", x="Blood Pressure (mmHg)", y="Frequency") +
  theme(panel.background = element_rect(fill = "white"),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                axis.line = element_line(colour = "black"))

```

```{r}
library(car)
library(ggplot2)

# Fit a linear regression model of log(BPSysAve) against another variable, e.g. Age
fit <- lm(log(BPSysAve) ~ Age, data=final_data)

# Create a QQ plot of the residuals of the linear regression model
qq <- qqPlot(resid(fit), main="Normal Q-Q", xlab="Theoretical quantities", ylab="Standardized residuals")

# Modify the theme of the plot to set the background to white and remove the frame
qq + theme(panel.background = element_rect(fill = "white"),
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           axis.line = element_line(color = "white"))
```


```{r}
sd_of_Age <- sd(final_data$Age)
print(sd_of_Age)

sd_of_BMI <- sd(final_data$BMI, na.rm = TRUE)
print(sd_of_BMI)
```


Visualize the blood pressure difference in sex results in a histogram
```{r}
# load the ggplot2 package
library(ggplot2)

# calculate the mean BPSysAve for males and females
male_mean <- mean(final_data$BPSysAve[final_data$Gender == "male"], na.rm = TRUE)
female_mean <- mean(final_data$BPSysAve[final_data$Gender == "female"], na.rm = TRUE)

# create a data frame for the mean BPSysAve values
mean_sex_data <- data.frame(Gender = c("Male", "Female"),
                        Mean_BPSysAve = c(male_mean, female_mean))

# create the plot with the gray color scheme
sex_bar <- ggplot(mean_sex_data, aes(x = Gender, y = Mean_BPSysAve, fill = Gender)) + 
          geom_bar(stat = "identity", color = "white", fill = "lightgray", width = 0.5) +
          geom_text(aes(label = round(Mean_BPSysAve, 1)), vjust = -0.5) +
          labs(x = "Gender", y = "Mean Systolic Blood Pressure (mmHg)", title = "Mean systolic blood pressure difference by gender") +
          theme(panel.background = element_rect(fill = "white"),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                axis.line = element_line(colour = "black"))

# display the plot
sex_bar


```

Visualize the blood pressure difference in race results in a histogram
```{r}
# calculate the mean BPSysAve for races
Black_mean <- mean(final_data$BPSysAve[final_data$Race1 == "Black"], na.rm = TRUE)
Hispanic_mean <- mean(final_data$BPSysAve[final_data$Race1 == "Hispanic"], na.rm = TRUE)
Mexican_mean <- mean(final_data$BPSysAve[final_data$Race1 == "Mexican"], na.rm = TRUE)
White_mean <- mean(final_data$BPSysAve[final_data$Race1 == "White"], na.rm = TRUE)
Other_mean <- mean(final_data$BPSysAve[final_data$Race1 == "Other"], na.rm = TRUE)


# create a data frame for the mean BPSysAve values
mean_race_data <- data.frame(Race1 = c("Black", "Hispanic", "Mexican", "White", "Other"),
                        Mean_BPSysAve = c(Black_mean, Hispanic_mean, Mexican_mean, White_mean, Other_mean))

# create the plot with the gray color scheme
race_bar <- ggplot(mean_race_data, aes(x = Race1, y = Mean_BPSysAve, fill = Race1)) + 
          geom_bar(stat = "identity", color = "white", fill = "lightgray", width = 0.5) +
          geom_text(aes(label = round(Mean_BPSysAve, 1)), vjust = -0.5) +
          labs(x = "Race", y = "Mean Systolic Blood Pressure (mmHg)", title = "Mean systolic blood pressure difference by race") +
          theme(panel.background = element_rect(fill = "white"),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                axis.line = element_line(colour = "black"))

# display the plot
race_bar
```


Visualize the blood pressure difference in education results in a histogram
```{r}
# calculate the mean BPSysAve for education levels
eight_mean <- mean(final_data$BPSysAve[final_data$Education == "8th Grade"], na.rm = TRUE)
nine_eleven_mean <- mean(final_data$BPSysAve[final_data$Education == "9 - 11th Grade"], na.rm = TRUE)
high_school_mean <- mean(final_data$BPSysAve[final_data$Education == "High School"], na.rm = TRUE)
some_college_mean <- mean(final_data$BPSysAve[final_data$Education == "Some College"], na.rm = TRUE)
college_grad_mean <- mean(final_data$BPSysAve[final_data$Education == "College Grad"], na.rm = TRUE)


# create a data frame for the mean BPSysAve values
mean_Education_data <- data.frame(Education = c("8th Grade", "9 - 11th Grade", "High School", "Some College", "College Grad"),
                        Mean_BPSysAve = c(eight_mean, nine_eleven_mean, high_school_mean, some_college_mean, college_grad_mean))

# create the plot with the gray color scheme
Education_bar <- ggplot(mean_Education_data, aes(x = Education, y = Mean_BPSysAve, fill = Education)) + 
          geom_bar(stat = "identity", color = "white", fill = "lightgray", width = 0.5) +
          geom_text(aes(label = round(Mean_BPSysAve, 1)), vjust = -0.5) +
          labs(x = "Education level", y = "Mean Systolic Blood Pressure (mmHg)", title = "Mean systolic blood pressure difference by education level") +
          theme(panel.background = element_rect(fill = "white"),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                axis.line = element_line(colour = "black"))

# display the plot
Education_bar
```


Visualize the blood pressure difference in income results in a histogram
```{r}
# calculate the mean BPSysAve for income levels 
two_mean <- mean(final_data$BPSysAve[final_data$HHIncome == "25000-34999"], na.rm = TRUE)
three_mean <- mean(final_data$BPSysAve[final_data$HHIncome == "35000-44999"], na.rm = TRUE)
four_mean <- mean(final_data$BPSysAve[final_data$HHIncome == "45000-54999"], na.rm = TRUE)
seven_mean <- mean(final_data$BPSysAve[final_data$HHIncome == "75000-99999"], na.rm = TRUE)
more_mean <- mean(final_data$BPSysAve[final_data$HHIncome == "more 99999"], na.rm = TRUE)


# create a data frame for the mean BPSysAve values
mean_income_data <- data.frame(HHIncome = c("25000-34999", "35000-44999", "45000-54999","75000-99999", "more 99999"),
                        Mean_BPSysAve = c(two_mean, three_mean, four_mean,seven_mean, more_mean))

# create the plot with the gray color scheme
income_bar <- ggplot(mean_income_data, aes(x = HHIncome, y = Mean_BPSysAve, fill = HHIncome)) + 
          geom_bar(stat = "identity", color = "white", fill = "lightgray", width = 0.5) +
          geom_text(aes(label = round(Mean_BPSysAve, 1)), vjust = -0.5) +
          labs(x = "Annual income($)", y = "Mean Systolic Blood Pressure (mmHg)", title = "Mean systolic blood pressure difference by income") +
          theme(panel.background = element_rect(fill = "white"),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                axis.line = element_line(colour = "black"))

# display the plot
income_bar
```


Visualize the blood pressure difference in marital status in a histogram
```{r}
# calculate the mean BPSysAve for males and females
Divorced_mean <- mean(final_data$BPSysAve[final_data$MaritalStatus == "Divorced"], na.rm = TRUE)
LivePartner_mean <- mean(final_data$BPSysAve[final_data$MaritalStatus == "LivePartner"], na.rm = TRUE)
Married_mean <- mean(final_data$BPSysAve[final_data$MaritalStatus == "Married"], na.rm = TRUE)
NeverMarried_mean <- mean(final_data$BPSysAve[final_data$MaritalStatus == "NeverMarried"], na.rm = TRUE)
Separated_mean <- mean(final_data$BPSysAve[final_data$MaritalStatus == "Separated"], na.rm = TRUE)
Widowed_mean <- mean(final_data$BPSysAve[final_data$MaritalStatus == "Widowed"], na.rm = TRUE)


# create a data frame for the mean BPSysAve values
mean_marital_data <- data.frame(MaritalStatus = c("Divorced", "LivePartner", "Married","NeverMarried", "Separated", "Widowed"),
                        Mean_BPSysAve = c(Divorced_mean, LivePartner_mean, Married_mean,NeverMarried_mean, Separated_mean, Widowed_mean))

# create the plot with the gray color scheme
marital_bar <- ggplot(mean_marital_data, aes(x = MaritalStatus, y = Mean_BPSysAve, fill = MaritalStatus)) + 
          geom_bar(stat = "identity", color = "white", fill = "lightgray", width = 0.5) +
          geom_text(aes(label = round(Mean_BPSysAve, 1)), vjust = -0.5) +
          labs(x = "Marital status", y = "Mean Systolic Blood Pressure (mmHg)", title = "Mean systolic blood pressure difference by marital status") +
          theme(panel.background = element_rect(fill = "white"),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                axis.line = element_line(colour = "black"))

# display the plot
marital_bar
```


aim 1: 
```{r}
t.test(final_data$BPSysAve[final_data$Gender=="male"],final_data$BPSysAve[final_data$Gender=="female"])
```

```{r}
# Subset the data by gender
male_bpsys <- final_data$BPSysAve[final_data$Gender == "male"]
female_bpsys <- final_data$BPSysAve[final_data$Gender == "female"]

# Perform t-test
ttest <- t.test(male_bpsys, female_bpsys)

# Create box plot
boxplot(male_bpsys, female_bpsys, names = c("Male", "Female"), ylab = "BPSysAve")

# Add text to the plot to show the t-test results
text(x = 1.5, y = max(c(male_bpsys, female_bpsys)) - 10, 
     labels = paste("p-value = ", signif(ttest$p.value, 3)), pos = 3)

```


```{r}
summary(final_data)
```

```{r}
# load necessary libraries 
library(tidyverse)

# Conduct linear regression with alcohol consumption and systolic blood pressure
model <- lm(BPSysAve ~ AlcoholYear, data = final_data)

# Print model summary
summary(model)

# Test whether regression coefficient for alcohol consumption is significantly different from zero
coef_test <- coef(summary(model))["AlcoholYear", "Pr(>|t|)"]
if (coef_test < 0.05) {
  print("The regression coefficient for alcohol consumption is significantly different from zero.")
} else {
  print("There is no significant association between alcohol consumption and systolic blood pressure.")
}


```


```{r}
# Conduct linear regression with alcohol consumption and systolic blood pressure
model1 <- lm(BPSysAve ~ AlcoholYear + Gender + Age + BMI + MaritalStatus + HHIncome +Education + Race1, data = final_data)

# Print model summary
summary(model1)
```

```{r}
# Extract the coefficients and standard errors from the model fit summary
coefs <- coef(model1)
se <- sqrt(diag(vcov(model1)))

print(coefs)
print(se)
```

```{r}
# calculate the CIs 
ci <- confint(model1)
print(ci)
```


```{r}
estimates_df <- data.frame(predictor = names(coefs),
                    estimates = coefs,
                    lower = ci[, 1],
                    upper = ci[, 2])

library(dplyr)

# Create a named vector of new labels
estimates_df$predictor[estimates_df$predictor == "Gendermale"] <- "Gender: male"
estimates_df$predictor[estimates_df$predictor == "AlcoholYear"] <- "Alcohol consumption year"
estimates_df$predictor[estimates_df$predictor == "MaritalStatusMarried"] <- "Marital status: married"
estimates_df$predictor[estimates_df$predictor == "MaritalStatusNeverMarried"] <- "Marital status: never married"
estimates_df$predictor[estimates_df$predictor == "MaritalStatusDivorced"] <- "Marital status: divorced"
estimates_df$predictor[estimates_df$predictor == "Race1Hispanic"] <- "Race: Hispanic"
estimates_df$predictor[estimates_df$predictor == "Race1White"] <- "Race: White"
estimates_df$predictor[estimates_df$predictor == "Race1Other"] <- "Race: Othe"
estimates_df$predictor[estimates_df$predictor == "Race1Mexican"] <- "Race: Mexican"
estimates_df$predictor[estimates_df$predictor == "Education9 - 11th Grade"] <- "Education: 9-11th grade"
estimates_df$predictor[estimates_df$predictor == "Education8th Grade"] <- "Education: 8th grade"
estimates_df$predictor[estimates_df$predictor == "EducationHigh School"] <- "Education: High school"
estimates_df$predictor[estimates_df$predictor == "EducationSome College"] <- "Education: Some college"
estimates_df$predictor[estimates_df$predictor == "EducationCollege Grad"] <- "Education: College grad"
estimates_df$predictor[estimates_df$predictor == "HHIncome35000-44999"] <- "Income: 35 -< 45k"
estimates_df$predictor[estimates_df$predictor == "HHIncome45000-54999"] <- "Income: 45 -< 55k"
estimates_df$predictor[estimates_df$predictor == "HHIncome55000-64999"] <- "Income: 55 -< 65k"
estimates_df$predictor[estimates_df$predictor == "HHIncome65000-74999"] <- "Income: 65 -< 75k"
estimates_df$predictor[estimates_df$predictor == "HHIncome75000-99999"] <- "Income: 75 -< 100k"
estimates_df$predictor[estimates_df$predictor == "HHIncomemore 99999"] <- "Income: >- 100K"

print(estimates_df)

estimates_df <- estimates_df[-c(5,8,9,10,11,12,13,14), ]

print(estimates_df)

```



```{r}
ggplot(estimates_df, aes(y = predictor, x = estimates, xmin = lower, xmax = upper)) +
  geom_pointrange() +
  labs(x = "Estimates (95% CI)", y = "") + 
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  theme_classic() +
  annotate("text", x = .8, y = 6.4, label = "") + 
  annotate("text", x = 1.2, y = 6.4, label = "") +
  annotate("text", x = 2, y = 6.8, label = "")
```






```{r}
# Create a new column in the dataframe to store the sign of the estimates
estimates_df$sign <- ifelse(estimates_df$estimates < 0, "negative", "positive")
# Define colors for negative and positive estimates
colors <- c("negative" = "red", "positive" = "blue")
# Create the plot with color-coded lines
ggplot(estimates_df, aes(y = predictor, x = estimates, xmin = lower, xmax = upper, color = sign)) +
  geom_pointrange() +
  geom_linerange(aes(x = lower, xend = upper), size = 1) +
  labs(x = "Estimates (95% CI)", y = "") + 
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5, color = "blue") +
  scale_color_manual(values = colors) +
  theme_classic()

```



scatter plot
```{r}
# fit the linear model
model2 <- lm(BPSysAve ~ AlcoholYear + Age + Gender + BMI, data = final_data)

df <- data.frame(fitted.values = model2$fitted.values, residuals = model2$residuals)
print(df)
# create a new empty plot
plot.new()
# Create the residuals vs fitted values plot
plot(residuals ~ fitted.values, data = df, type = "p", xlab = "Fitted Values", ylab = "Residuals", main = "Residuals vs Fitted Values")
# Add a reference line
abline(h = 0, lty = 2, col = "red")


```






